Active Code Completion

### 2012-03-21 icse12/graphite-icse12.tex
Final ICSE2012 paper.

\begin{abstract}
Code completion menus have replaced standalone API browsers for most developers because they are more tightly  integrated into the development workflow. Refinements to the code completion menu that incorporate additional sources of information have similarly been shown to be valuable, even relative to standalone counterparts offering similar functionality. In this paper, we describe {\em active code completion}, a technique that enables the integration of a much broader array of interactive and highly-specialized code generation interfaces, called {\em palettes}, directly into the editor environment. Using several empirical methods, we examine the contexts in which such a system could be useful, describe the design constraints governing the system architecture as well as particular code completion interfaces, and design one such system, named Graphite, for the Eclipse Java development environment. Using Graphite, we implement a palette for writing regular expressions as our primary  example and conduct a small pilot study that provides further evidence in support of the claim that integrating specialized tools directly into the editor environment is both feasible, particularly if aided by an active code completion system like Graphite, and valuable to professional developers.
\end{abstract}

@@

Several refinements and additions to the code completion menu have previously been suggested in the literature. These have focused on leveraging additional sources of information, such as databases of usage history \cite{robbes_how_2008}\cite{HouPletcher2011}, inheritance information \cite{HouPletcher2011}, API-specific information \cite{HouPletcher2011}\cite{Lee+2008}, partial abbreviations \cite{Han+2009}, examples extracted from code repositories \cite{bruch_learning_2009}\cite{Brandt+2010} and crowdsourced information \cite{mooty_calcite:_2010}\cite{SnipMatch}, to increase the relevance and sophistication of the featured menu items. As with the standard form of code completion, many of these sources of data can also be utilized via external tools (e.g. Calcite \cite{mooty_calcite:_2010} uses information that could already be accessed using the Jadeite \cite{conf/vl/StylosFYM09} tool). Empirical evidence presented in these studies, however, indicates that directly integrating these kinds of tools directly into the editor environment can be more effective.

In all of these systems, the code completion interface has remained primarily menu-based. When an item is selected, code is inserted immediately, without further input from the developer. These systems are also limited in that they are difficult to modularly extend. A fixed strategy determines the completions that are available, leaving library providers unable to specify new domain-specific or contextually-relevant items. 
In this paper we propose a technique called {\it active code completion} that eliminates these restrictions \cite{ACC_VLHCC}. This  makes developing and integrating a broad array of highly-specialized developer tools directly into the editor, via the familiar code completion command, significantly simpler. This technique is motivated by the evidence discussed above and further evidence provided in this paper that developers prefer, and make more effective use of, tools that do not require leaving the immediate editing environment.

@@

If the developer invokes the code completion command at the indicated cursor position, the editor  looks  for a {\it palette definition} associated with the {\it type} of the expression being entered, which in this case is  \verb|Color|. If an associated palette is found, a menu item briefly describing this palette is added to the standard code completion menu. When selected, the corresponding palette is shown, replacing the standard code completion menu. Figure 1a gives an example of a simple palette that may be associated with the \verb|Color| class.

The developer can interact with such palettes to provide parameters and other information related to her intent, and receive immediate feedback about the effect these choices will have on the behavior of the object being constructed. When this interaction is complete, the palette generates appropriate source code for insertion at the cursor. Figure 1b shows the inserted code after the user presses the \textsc{enter}.

In accordance with rational design practices, we sought to address the following questions before designing and implementing our active code completion system:

@@

To help us answer these questions, we conducted a large online survey of professional developers (Section 2). Their responses, along with information gathered from informal interviews and code corpus analyses, revealed a number of interesting use cases (Section 3) and non-trivial functional requirements for palette interfaces as well as the underlying active code completion architecture (Section 4). Notably, many of these findings can be applied to development tools for professional developers more generally, not simply those integrated into the editor in the way we have shown.

Next, we describe Graphite, an Eclipse plug-in that provides active code completion for the Java programming language (Section 5). We describe several design choices that were made to satisfy many of the requirements discovered in our preliminary investigations and briefly examine the trade-offs that were necessary due to Java's purely textual program representations and Eclipse's limited extension APIs.

Finally, we conducted a pilot lab study with a more complex palette, implemented using Graphite, that assists developers as they write regular expressions (Section 6). The data and observations gathered from this study provide specific evidence in support of the broader claim that highly-specialized tools that are integrated directly with the editing environment are particularly useful, and that an active code completion system like Graphite makes developing, deploying and discovering such tools fundamentally simpler.

\section{Survey}
To validate our general conceptualization of active code completion, develop concrete design criteria and create a concrete list of use cases, we began by conducting a large survey of professional software developers.

% In order to convey the active code completion concept clearly, we came up with three different use cases where we belive the active code completion technique will benefit developers frequently. For each use case, we designed three mockup screenshots and showed them in sequence: before invoking the palette, during the interaction with the palette, and after the palette is closed and code is generated. By doing this, the respondents could understand the concept precisely and give useful feedback relevant to the above questions.
\subsection{Survey Methods}
We recruit participants for this survey\footnote{https://www.surveymonkey.com/s/2GLZP8V} primarily from a popular programming-related discussion forum hosted on the popular website reddit.com \cite{reddit_programming}. An additional 22 participants were recruited from the local computer science graduate student body.

@@

Before asking participants about regular expressions and SQL queries, we asked participants to rate their familiarity with these concepts. Few participants (4\%) indicated that they were unfamiliar with regular expressions and no participants were unfamiliar with creating SQL queries, providing further evidence that our participants were not novice developers. Fig. \ref{strategies} summarizes the strategies that participants generally preferred for instantiating regular expressions and SQL queries. Using an external tool was a common strategy in both cases, particularly for SQL queries, but several other strategies were also represented.

@@

Finally, after showing the series of mockup screenshots, we asked participants to rate how useful the integrated palette would be to them if they needed to instantiate the corresponding class. The responses to this question for each palette are summarized in Fig. \ref{frequencies}. In each case, more than half of the participants indicated that they would use active code completion at least some of the time. The regular expression palette was considered particularly useful while the color and SQL palettes showed a more reserved pattern of responses.

@@

\section{Design Criteria}
Using the information gathered from the survey as well as informal discussions with developers and researchers over the course of this study, we sought to develop design criteria constraining both the overall system design as well as the design of individual palettes. In the section headings below, the number of survey responses, summed over the three palette mockups, that contained the listed concern, as judged by the authors of this paper, are listed in parenthesis. Many of these concerns apply generally to various kinds of editor-integrated tools, not just those implemented using the active code completion mechanism that we focus on, so we hope these design criteria will be valuable to the developer tool design community more broadly.

\subsection{Maintaining Separation of Concerns (183)}
The most common issue developers had, particularly with the Color palette, was that it ignored the issue of separation of concerns by allowing developers to insert color constants directly into the program logic. Many developers noted that this is considered bad practice, or should be limited to the prototyping phase of a project. This concern was also expressed in responses to the SQL palette, which required inserting parameters to connect to a database so that the query could be tested. The resulting code included initialization steps needed to connect to the specific database that was entered, and several participants noted that much of this information should appear in an external resource file separated from the program logic. Few participants made similar comments about the regular expression palette, however, indicating that regular expressions are considered a part of the program logic rather than data by most developers.

This feedback suggests that tool and palette designers may wish to explicitly aprise users of relevant best practices and acknowledge that palettes that generate constant data may be most useful in the prototyping phase. It can be noted that when transitioning from a prototype to production-quality software the code generated by a palette or tool may be used as template that can be refactored to follow stricter policies as needed. We showed users the Color and SQL palettes in the context of a standard programming editor, but these comments indicate that resource file and  stylesheet editors may benefit from active code completion support as well.

\subsection{Integration with Testing Frameworks (35)}
The regular expression palette shown to the participants allowed users to immediately test a pattern against provided strings. These test strings and the results of performing the match were inserted as comments below the generated source code. A number of participants requested that explicit unit tests be generated instead. Although the palette interface conveyed the results of these tests immediately, participants remained adamant about explicit unit tests, likely due to concerns that future modifications to the generated code without reinvoking the palette might introduce bugs, or due to the desire to conform to standard testing practices. To support the generation of unit tests, however, the active code completion architecture must explicitly allow for code generation at locations other than at the cursor and handle the fact that statements generated after a code completion command are not necessarily encapsulated well enough to be immediately testable in isolation.

@@

A related feature important to many participants is support for maintaining settings and usage history across invocations of the same palette at different code locations. For example, 20 participants requested that the Color palette include a list of recent or favorite colors, and 12 participants specifically inquired about whether the database connection information was maintained between invocations of the SQL palette.

@@

\subsection{Minimal Impact on Editor Responsiveness}

@@

Our initial color palette was deemed overly complex by many participants (26), with many others (26) specifically wanting a simpler interface with colors listed next to their names. Other users wanted additional features, such as an ``eyedropper'' tool for selecting a color directly from an image on the screen. Similarly, the regular expression and SQL palettes were considered too simple by 12, and 15 participants respectively. User requests included syntax highlighting, a range of additional helpful tools for generating, testing and sharing regular expressions, and the ability to browse databases more directly. Due to the wide range of user needs even within a single class, it may be that support for multiple palettes or a tabbed palette interface will be necessary in practice. To support incremental improvements based on user feedback, an architecture that makes deploying new and updated palettes relatively painless would also be valuable.

\section{Use Cases}
At the end of the survey, we solicited suggestions from the participants of other classes that could benefit from an associated palette. 119 participants made one or more suggestion and we classified the suggestions into several categories. We also mention other suggestions made by researchers and developers via other channels in some cases, without including them in the provided counts.

@@

\subsection{Proof Assistants (1)}
A proof assistant is a tool for constructing proof terms. According to the well-known Curry-Howard isomorphism between programming languages and formal logics, proof terms correspond to expressions and propositions correspond to types \cite{tapl}. Active code completion works directly with types to help developers construct expressions, so if applied to a language with a cleanly-developed connection to formal logic (e.g. Coq), palettes would be useful for constructing highly interactive proof assistant interfaces.

@@

\section{Tool Design}
Motivated by these use cases and with the design criteria developed as a result of our developer survey in mind, we sought to create an active code completion system. We named our project Graphite, an acronym for {\bf Gra}phical {\bf P}alettes to {\bf H}elp {\bf I}nstantiate {\bf T}ypes in the {\bf E}ditor. We describe the design decisions and trade-offs that went into this system in this section. A video demonstrating Graphite, including the process of writing a full palette, is available for further details \footnote{http://www.youtube.com/watch?v=FfCoXQEHGAI}. The full source code of Graphite is also available at the URL given on the first page.

We chose to use Eclipse for the Java programming language because this combination was the most widely-used  amongst participants in our survey. Eclipse plug-in development is well-supported and there are many examples available online that we can refer to, and the Eclipse project is available under an open source license.

\subsection{HTML+Javascript based palettes}
Eclipse is itself written in Java and uses the SWT and JFace graphical user interface toolkits. These frameworks are not widely used outside of the Eclipse ecosystem. As we have noted, a number of participants in our field study indicated that they hoped that our tool would be available for other IDEs and other programming languages. Indeed, many concepts (including regular expressions, colors and database queries) are similar across languages. To better support IDE-independence and language-independence, we chose instead to build palettes using HTML and Javascript. Javascript was among the most well-known languages in our study, just behind Java, C and C++, and is highly flexible. A number of user interface libraries are available, such as jQuery\footnote{http://jquery.com/} (which we used to implement our palettes). All major browsers now feature sophisticated debuggers and run-time inspection facilities that ease debugging and testing. Deploying palettes using standard URLs is also simpler than attempting to integrate their source code directly into Java libraries and packages and eases the process of incremental development and deployment. All major windowing toolkits feature a web browser control that can allow for the display of these palettes within the editor environment. In Eclipse, we used the standard SWT Browser control.

\subsection{Palette API}
In order to allow palettes to communicate with the surrounding editor environment, a simple Javascript interface was injected. These functions were implemented by the editor plug-in, but appear as ``native'' browser functions to Javascript code. The API consists of the following methods, accessed through a global object named \verb|graphite|:
\begin{itemize}
	\item \verb|insert(str)|: Inserts the specified string at the cursor and closes the palette. The indentation is automatically inserted after any newlines in this string.

	\item \verb|cancel()|: Closes the palette without changes.

	\item \verb|getSelectedText()|: Returns the text that is currently selected by the user in the editor, or an empty string if no text has been selected. This method is used to implement reinvocation.

	\item \verb|getIDE()|: Returns a string that specifies the IDE that is being used, ``Eclipse'' in our implementation.

	\item \verb|getLanguage()|: Returns a string that specifies the programming language that is being used, ``java'' in our implementation.
\end{itemize}

By limiting the complexity of this API, we reasoned that developers would be able to create specialized palettes more easily. An additional benefit is that the developers of other editing environments are also able to create plug-ins that support Graphite palettes with minimal effort.

To access this API, the palette must include a small script named \verb|graphite.js| into their page. This file interacts with the host editor to create the graphite object and provides an alternative implementation of these methods when the palette is being tested in a web browser.

\subsection{Palette Discovery}
\begin{figure*}\begin{center}
\includegraphics[scale=.6]{regex.png}\end{center}
\caption{The regular expression palette developed using Graphite and used by subjects in the treatment group of the user study described in Sec. VI.}
\end{figure*}

Graphite currently provides two methods for associating a palette with a particular type so that the Graphite plug-in is able to discover it when the code completion menu is invoked by a developer.

\subsubsection{Annotation-based}
For user-defined classes that the palette author has the authority to modify, the \texttt{@GraphitePalette} annotation associates a palette with the class. The annotation must specify the URL of the palette, and can contain some other optional information (e.g. the description that is shown in the code completion menu). This  allows library providers to provide palettes that are specialized to their libraries and distribute them alongside their code. In this way, the users of the libraries are not required to discover that an external tool exists. If the user does not have the Graphite plug-in for Eclipse installed, the annotation has no effect.

\subsubsection{Explicit}
In cases where the palette developer cannot modify a class directly (such as palettes for classes in the Java standard library), a second method exists to explicitly associate a palette with the fully qualified name of a class. Currently, this feature is exposed as a preference pane in the Eclipse IDE. To facilitate automated installs in the future, a preferences file may also be added to the system.

\subsection{Palette Reinvocation}
To make further changes on the code generated by a Graphite palette, users can select the generated code and reinvoke the palette. The individual palette is responsible for parsing the selected code and filling the parameters back in the palette appropriately. When the user is satisfied with the new parameters and hits {\tt ENTER}, the selected code is replaced with a newly generated code.

\subsection{Design Trade-Offs}
The design that we have described is light-weight, highly flexible and highly responsive. It lays foundations for IDE and language portability. However, several trade-offs have also been made to achieve these benefits.

\begin{itemize}
\item Because palettes are implemented in Javascript, any differences between Java and Javascript become evident. For example, the color names are slightly different between Java and Javascript, and so are their regular expression engines. Although a Java applet could be used in cases where these differences are critical, this is still more difficult than it would be if the palettes were implemented using the same language as the code.
\item The palette user interface must stay within its bounding box -- user interface elements like pop-up menus (unless they are provided by the browser itself) are thus more difficult to implement and may require additional API support in the future.
\item Several use cases could benefit from greater access to the surrounding code, or even the surrounding project. Implementing this modularly is particularly difficult, and the Eclipse API does not easily allow for  serialization of the sum of its contextual knowledge for consumption by a palette.
\item Implementing a robust reinvocation mechanism that does not require the user to select the generated code is difficult, because this would require the palette to parse all the code to find relevant information, and also remove the ``dead'' code. This puts a greater burden on palette developers and errors could affect user's code in subtle ways. The small inconvenience of having to select code before reinvoking a palette may be balanced by the knowledge that no other portions of the code will be modified inadvertently by that palette.
\end{itemize}

\subsection{Palettes}
We implemented two full-featured palettes to demonstrate and test the functionality of this framework: a regular expression palette (Figure 5) and a color selection palette (Figure 1).

To allow users to test the behavior of the regular expression that they have entered, two columns are available below the expression input area. The left column contains an input area labeled ``Should match...'' and the right column contains an input area labeled ``Should NOT match...''. Users use these to enter lists of test  strings into each column. The background colors behind these strings change to indicate whether the regular expression that has been entered matches or does not match that string. Green indicates that the pattern  {\it matched} the string and red indicates that it does {\it not} match, regardless of the column that the test is in (this scheme was chosen based on feedback from an initial pilot study of this palette.) A key describing this color scheme is displayed after the first test has been entered (not shown).

Users navigate between the three input areas using standard \verb|Tab| cycling behavior. The label in each text area remains visible until some input has been entered, rather than disappearing immediately on focus. When a user is satisfied with the regular expression that they have entered, she can press the \verb|Enter| key to insert the appropriate Java source code. Because Java requires that the regular expression pattern be placed inside a string literal, additional escape sequences are needed in front of backslashes. This can be tedious and error-prone if done manually. Because the palette is responsible for generating this literal, however, this is no longer a problem. In addition to the source code itself, the tests are retained in a comment beginning on the next line. The user can leave these as documentation or use them to create more formal unit tests within their preferred testing framework.

If the user wishes to modify the regular expression or change the test set, she can highlight the code that was inserted and then invoke the palette once again. The palette parses the selected text to extract the regular expression and tests. Although this is slightly inconvenient in that it requires that the user select the relevant portion of the code, it is the most robust method for implementing this feature -- it can easily handle changes to the regular expression or tests and it does not require additional logic in the palette to determine which portions of the code are relevant, and should subsequently be replaced if a new regular expression is entered.

\subsubsection{Color Selection}
We omit a description of the full behavior of the Color selection palette due to space limitations. 
%\section{Future work}
%
%\subsection{Graphite plug-in for different IDEs}
%As all the palettes are independent to Eclipse, We speculate that implementing Graphite plug-in for other IDEs will not be too difficult. Implementing a web-based aiding tool and sharing that tool among different IDEs with minimal effort will be a good idea.
%
%\subsection{Improving the Graphite API for Javascript}
%The current version of Graphite API for Javascript is not powerful enough. For example, the palettes cannot add comments before current line, nor can it delete or replace code fragment which was not highlighted by the user. By providing more effective and powerful API, the palette writers can write the palettes more quickly and make more useful palettes.
%
%\subsection{More palette examples}
%Due to the limited time, we could only provide two example palettes so far. More palette examples for various types might inspire the library providers and palette writers to design better palettes.
%
%\subsection{Keyword invocation model / online palette repository}
%Although Keyword invocation has many expected benefits, it is currently not supported. This model will make Graphite language-independent, and improve discoverability of the palettes by allowing users to search the palette with keywords. Palette search will work better if a shared online palette repository is built and maintained.
%
%

@@

\section{User Study Design}

We conducted a comparative study to evaluate the usefulness and usability of a Graphite palette in a constrained setting -- writing regular expressions. We used a between-subjects design by randomly splitting the participants into two groups: a control group, where subjects were not shown nor able to use the palette, and a treatment group, where subjects were shown a different palette and allowed to discover and use the regular expression palette if they wished.

\subsection{Process}
\subsubsection{Preliminary Survey}
The participants were first asked to fill out the pre-survey form about their prior experiences with various programming languages, integrated development environments (IDEs), and regular expressions, based on the previous survey that we had conducted. Of note, the subjects in this study were considerably less skilled with Java and regular expressions than in the previous survey, likely due to a sampling effect -- our subjects were all PhD students at Carnegie Mellon.

>>>>

\begin{abstract}
Code completion menus have replaced standalone API browsers for most developers because they are more tightly  integrated into the development workflow. Refinements to the code completion menu that incorporate additional sources of information have similarly been shown to be valuable, even relative to standalone counterparts offering similar functionality. In this paper, we describe {\em active code completion}, an architecture that allows library developers to introduce interactive and highly-specialized code generation interfaces, called {\em palettes}, directly into the editor. Using several empirical methods, we examine the contexts in which such a system could be useful, describe the design constraints governing the system architecture as well as particular code completion interfaces, and design one such system, named Graphite, for the Eclipse Java development environment. Using Graphite, we implement a palette for writing regular expressions as our primary  example and conduct a small pilot study. In addition to showing the feasibility of this approach, it provides further evidence in support of the claim that integrating specialized code completion interfaces directly into the editor is valuable to professional developers.
\end{abstract}

@@

Several refinements and additions to the code completion menu have previously been suggested in the literature. These have focused on leveraging additional sources of information, such as databases of usage history \cite{robbes_how_2008}\cite{HouPletcher2011}, inheritance information \cite{HouPletcher2011}, API-specific information \cite{HouPletcher2011}\cite{Lee+2008}, partial abbreviations \cite{Han+2009}, examples extracted from code repositories \cite{bruch_learning_2009}\cite{Brandt+2010} and crowdsourced information \cite{mooty_calcite:_2010}\cite{SnipMatch}, to increase the relevance and sophistication of the featured menu items. As with the standard form of code completion, many of these sources of data can also be utilized via external tools (e.g. Calcite \cite{mooty_calcite:_2010} uses information that could already be accessed using the Jadeite \cite{conf/vl/StylosFYM09} tool). Empirical evidence presented in these studies, however, suggests that directly integrating these kinds of tools into the editor is particularly effective. For example, users of the Calcite tool completed 40\% more tasks in a lab study (unfortunately, a Jadeite control group was not included.)

In all of these systems, the code completion interface has remained primarily menu-based. When an item is selected, code is inserted immediately, without further input from the developer. These systems are also difficult to extend: a fixed strategy determines the completions that are available, so library providers cannot directly specify new domain-specific or contextually-relevant logic. 
In this paper we propose a technique called {\it active code completion} that eliminates these restrictions\footnote{Portions of this work previously appeared in a poster abstract \cite{ACC_VLHCC}.}. This  makes developing and integrating a broad array of highly-specialized developer tools directly into the editor, via the familiar code completion command, significantly simpler. This technique is motivated by the evidence discussed above and further evidence provided in this paper that developers prefer, and make more effective use of, tools that do not require leaving the immediate editing environment.

@@

If the developer invokes the code completion command at the indicated cursor position (\textvisiblespace), the editor  looks  for a {\it palette definition} associated with the {\it type} of the expression being entered, which in this case is  \verb|Color|. If an associated palette is found, a menu item briefly describing this palette is added to the standard code completion menu. When selected, the corresponding palette is shown, replacing the standard code completion menu. Figure 1a gives an example of a simple palette that may be associated with the \verb|Color| class\footnote{A video demonstrating this process is available from the webpage above.}.

The developer can interact with such palettes to provide parameters and other information related to her intent, and receive immediate feedback about the effect these choices will have on the behavior of the object being constructed. When this interaction is complete, the palette generates appropriate source code for insertion at the cursor. Figure 1b shows the inserted code after the user presses \textsc{enter}.

In accordance with best practices, we sought to address the following questions before designing and implementing our active code completion system:

@@

To help us answer these questions, we conducted a survey of 473 professional developers (Section II). Their responses, along with information gathered from informal interviews and code corpus analyses, revealed a number of non-trivial functional requirements for palette interfaces as well as the underlying active code completion architecture (Section III). Participants also suggested a large number of use cases, demonstrating the broad applicability of this technique. We organize these into several broad categories (Section IV).

Next, we describe Graphite, an Eclipse plug-in that implements the active code completion architecture for the Java programming language (Section V), allowing Java library developers to associate custom palettes with their own classes. We describe several design choices that we made to satisfy the requirements discovered in our preliminary investigations and briefly examine necessary trade-offs.

Finally, we conducted a pilot lab study with a more complex palette, implemented using Graphite, that assists developers as they write regular expressions (Section VI). The study provides specific evidence in support of the broader claim that highly-specialized tools that are integrated directly with the editing environment are particularly useful. We conclude that active code completion systems like Graphite are useful because they make developing, deploying and discovering such tools fundamentally simpler.

\section{Survey}
To validate our general conceptualization of active code completion, develop concrete criteria to constrain our system and palette designs, and create a list of use cases to justify this effort, we began by conducting a large survey of professional software developers.

% In order to convey the active code completion concept clearly, we came up with three different use cases where we belive the active code completion technique will benefit developers frequently. For each use case, we designed three mockup screenshots and showed them in sequence: before invoking the palette, during the interaction with the palette, and after the palette is closed and code is generated. By doing this, the respondents could understand the concept precisely and give useful feedback relevant to the above questions.
\subsection{Participants}
We recruited participants for this survey\footnote{https://www.surveymonkey.com/s/2GLZP8V} primarily from a popular programming-related discussion forum hosted on the popular website reddit.com \cite{reddit_programming}. An additional 22 participants were computer science graduate students at CMU.

@@

Before asking participants about regular expressions and SQL queries, we asked participants to rate their familiarity with these concepts. Few participants (4\%) indicated that they were unfamiliar with regular expressions and no participants were unfamiliar with creating SQL queries, providing further evidence that our participants were not novice developers. Figure \ref{strategies} summarizes the strategies that participants generally preferred for instantiating regular expressions and SQL queries. Using an external tool was a common strategy in both cases, particularly for SQL queries, but several other strategies were also represented.

@@

Finally, after showing the series of mockup screenshots, we asked participants to rate how useful the integrated palette would be to them if they needed to instantiate the corresponding class. The responses to this question for each palette are summarized in Figure \ref{frequencies}. In each case, more than half of the participants indicated that they would use active code completion at least some of the time. The regular expression palette was considered particularly useful while the color and SQL palettes showed a more reserved pattern of responses.

@@

\section{Design Criteria}
Using the information gathered from the survey as well as informal discussions with developers and researchers, we developed  design criteria constraining both the overall system design as well as the design of individual palettes. In the section headings below, the number of survey responses, summed over the three palette mockups, that contained the listed concern, as judged by the authors of this paper, are listed in parenthesis. These criteria were useful in designing Graphite (Section V) and we note that this collection of criteria may also be relevant to researchers designing other kinds of editor-integrated tools. Based on the variety of concerns expressed by participants in our survey, the design space for these tools appears to be quite complex.

\subsection{Maintaining Separation of Concerns (183)}
The most common issue participants had was that palettes seemed to violate the principle of separation of concerns. The color palette, for example, allows developers to insert color constants directly into the program logic. Many developers noted that this is considered bad practice, or should be limited to the prototyping phase of a project. This concern was also expressed in responses to the SQL palette, which required inserting parameters to connect to a database so that the query could be tested. The resulting code included initialization steps needed to connect to the specific database that was entered, and several participants noted that much of this information should appear in an external resource file separated from the program logic. Few participants made similar comments about the regular expression palette, however, indicating that regular expressions are considered a part of the program logic rather than data by most developers.

This suggests that tool and palette designers may wish to explicitly aprise users of relevant best practices and acknowledge that palettes that generate constant data may be most useful in the prototyping phase. It can be noted that when transitioning from a prototype to production-quality software the code generated by a palette or tool may be used as template to be refactored as needed. It also suggests that resource file and stylesheet editors may particularly benefit from active code completion support.

\subsection{Integration with Testing Frameworks (35)}
The regular expression palette shown to the participants allowed users to immediately test a pattern against provided strings. These test strings and the results of performing the match were inserted as comments below the generated source code. A number of participants requested that unit tests be generated instead, likely due to concerns that future modifications might introduce bugs, or due to the desire to conform to standard testing practices. To support the generation of unit tests, the active code completion architecture would need to support code generation at locations other than at the cursor.

@@

A related feature important to many participants was support for maintaining settings and usage history across invocations of the same palette at different code locations. For example, 20 participants requested that the Color palette include a list of recent or favorite colors, and 12 participants inquired about whether the database connection information was maintained between invocations of the SQL palette.

@@

\subsection{Responsiveness}

@@

While some participants wanted simpler palettes, others requested significant new capabilities, indicating that user needs may vary substantially. For example, our initial color palette (different from the one shown in Figure 1) was deemed overly complex by many participants (26). Other users wanted additional features, such as an ``eyedropper'' tool (11) for selecting a color directly from an image on the screen. The regular expression and SQL palettes were considered too simple (by 12, and 15 participants respectively). Users suggested syntax highlighting, several mechanisms for generating, testing and sharing regular expressions and queries, and the ability to browse SQL databases.

Due to this wide range of user needs even for a single class, it may be that support for multiple palettes or a tabbed palette interface would be helpful in practice. To support incremental improvements based on such user feedback, an architecture that makes deploying new and updated palettes relatively painless would also be valuable.

\section{Use Cases}
At the end of the survey, we asked the the participants to suggest other classes that could benefit from an associated palette to support our claim that active code completion is broadly applicable, and to allow us to characterize the specific scenarios where it may be most useful. A total of 119 participants made one or more suggestions, which we classified into several broad categories (we omit a few of these below due to space constraints). As above, the number of participants suggesting a palette in each category is listed in parenthesis. We also include suggestions made by researchers and developers in private discussions without including them in the provided counts.

@@

\subsection{Proof Assistants}
A proof assistant is a tool for constructing proof terms. According to the well-known Curry-Howard isomorphism between programming languages and formal logics, proof terms correspond to expressions and propositions correspond to types \cite{tapl}. Active code completion works directly with types to help developers construct expressions, so if applied to a language with a cleanly-developed connection to formal logic (e.g. Coq), palettes would be useful for constructing  interactive proof assistant interfaces.

@@

\section{System Design and Implementation}
After completing the survey, we built an active code completion system named Graphite, an acronym for {\bf Gra}phical {\bf P}alettes to {\bf H}elp {\bf I}nstantiate {\bf T}ypes in the {\bf E}ditor. 
We chose to build the system as an extension to Eclipse for Java because this combination was the most widely-used  amongst participants in our survey. In the subsections below, we describe how several novel design decisions made it possible to satisfy design criteria from Section III and enabled several use cases described in Section IV. The end result is a simple system that allows an API's developers, as well as external developers, to build rich HTML5-based palettes that can be associated with both in-built and user-defined classes directly. Developers using an API can discover and invoke these palettes through the standard code completion menu.

\subsection{HTML5-Based Palettes}
Palette developers build palettes using HTML5 technologies (HTML, CSS and JavaScript). We made this decision for several reasons. 
Eclipse is written in Java and uses the SWT and JFace graphical user interface toolkits, but these are not widely used outside of the Eclipse ecosystem. JavaScript was among the most well-known languages in our study, just behind Java, C and C++, and is highly flexible. A number of useful libraries are available (e.g. jQuery \cite{jQuery}.)

As noted in Section IV-H, a number of participants in our field study indicated that they hoped that our tool would be available for other IDEs and other programming languages. All major windowing toolkits feature a web browser control, so HTML5-based palettes can easily be loaded by different editor environments without developer intervention.

Deploying palettes using standard URLs is also simpler than attempting to integrate them directly into Java libraries and packages. It also eases the process of incremental and rapid development, as all major browsers now feature sophisticated debuggers and run-time inspection facilities.

\subsection{Palette API}
Palettes communicate with the host IDE using a simple Javascript API. To access this API, the palette must include a small script named \verb|graphite.js| into their page. The default implementation of this script provides an implementation of the API methods for testing in a standard web browser. When loaded into an editor, the editor plug-in replaces these methods with specialized implementations. The API consists of the following methods, accessed through a global object named \verb|graphite|:
\begin{itemize}
	\item \verb|insert(str)|: Inserts the specified string at the cursor and closes the palette. The indentation is automatically inserted after any newlines in this string.

	\item \verb|cancel()|: Closes the palette without inserting code. Note that the ESC key cancels by default, and palettes are modal so clicking outside a palette will not cancel.

	\item \verb|getSelectedText()|: Returns the text that is currently selected in the editor, or an empty string if no text has been selected. This method is used to implement reinvocation -- users simply select previously generated code and invoke the palette as described in Section I. The palette is responsible for parsing the selected text to extract relevant parameters.

	\item \verb|getIDE()|, \verb|getLanguage()|: Return a string that specifies the IDE and language that is being used, ``Eclipse'' and ``Java'' in our implementation.
\end{itemize}

By limiting the complexity of this API, we reasoned that developers would be able to create specialized palettes more easily. A simple hello world palette is only two lines of code, for example. An additional benefit is that the developers of other editing environments should also be able to create plug-ins that support Graphite palettes with minimal effort.

\subsection{Palette Discovery}
\begin{figure*}\begin{center}
\vspace{2mm}
\includegraphics[scale=.6]{regex.png}\end{center}
\caption{The regular expression palette developed using Graphite and used by subjects in the treatment group of the user study described in Sec. VI.}
\end{figure*}

Graphite currently provides two methods for associating a palette with a particular class so that the editor plug-in can include it in the code completion menu when relevant.

\subsubsection{Annotation-based}
For user-defined classes that the palette developer has the authority to modify, the \texttt{@GraphitePalette} annotation associates a palette with the class. The annotation must specify the URL of the palette, and can contain some other optional information (e.g. the description that is shown in the code completion menu). This  allows API providers to provide palettes that are specialized to their libraries and distribute them directly alongside their code. The benefit of this approach is that users of the API are not required to discover that an external tool exists and explicitly install it into their IDE.

\subsubsection{Explicit}
In cases where a palette developer cannot modify a class directly (such as palettes for classes in the Java standard library), end-users can explicitly associate a palette with the fully qualified name of a class via a preference pane in the Eclipse IDE.

\subsection{Design Trade-Offs}
The design that we have described is light-weight, highly flexible and does not significantly impact IDE responsiveness. It also lays foundations for IDE and language portability. However, this design also leads to trade-offs:

\begin{itemize}
\item Because palettes are implemented in Javascript, any differences between the semantics of Java and Javascript can be problematic. For example, color names are slightly different between Java and Javascript, as are the regular expression engines. Although a Java applet could be used in cases where these differences are critical, this is still more difficult than it would be if the palettes were implemented in the same language.
\item The palette user interface must stay within its bounding box -- user interface elements like pop-up menus (unless they are provided by the browser itself) are thus more difficult to implement and may require additional API support in the future.
\item Several use cases could benefit from greater access to the surrounding code, or even the surrounding project. Implementing this modularly is particularly difficult, and the Eclipse API does not easily allow for  serialization of the sum of its contextual knowledge for consumption by a palette.
\item Because the reinvocation mechanism relies on parsing the selected code, the burden is high for both palette developers (although Javascript libraries for parsing Java code are available) and palette users. A better solution would be one where palette-related metadata is stored directly with the generated code. This could be partially addressed by the use of special comments to delimit sections of generated code and store palette metadata, but a more elegant solution would require a  change in how Java source code is represented.
\end{itemize}

%\section{Future work}
%
%\subsection{Graphite plug-in for different IDEs}
%As all the palettes are independent to Eclipse, We speculate that implementing Graphite plug-in for other IDEs will not be too difficult. Implementing a web-based aiding tool and sharing that tool among different IDEs with minimal effort will be a good idea.
%
%\subsection{Improving the Graphite API for Javascript}
%The current version of Graphite API for Javascript is not powerful enough. For example, the palettes cannot add comments before current line, nor can it delete or replace code fragment which was not highlighted by the user. By providing more effective and powerful API, the palette writers can write the palettes more quickly and make more useful palettes.
%
%\subsection{More palette examples}
%Due to the limited time, we could only provide two example palettes so far. More palette examples for various types might inspire the library providers and palette writers to design better palettes.
%
%\subsection{Keyword invocation model / online palette repository}
%Although Keyword invocation has many expected benefits, it is currently not supported. This model will make Graphite language-independent, and improve discoverability of the palettes by allowing users to search the palette with keywords. Palette search will work better if a shared online palette repository is built and maintained.
%
%

@@

\subsection{Palettes}
We implemented two palettes using the jQuery library for basic functionality \cite{jQuery}, which we used in the pilot study described below.

\subsubsection{Color Selection}
The color selection palette we ultimately developed was significantly simpler than the one shown in the preliminary survey, due to user comments. It allows users to enter any valid CSS color string, satisfying the requirements for keyboard navigability. The entered color is syntax checked and a preview is shown. Standard Java colors are also available as swatches that can be selected by the mouse if desired. Reinvocation support is provided. This palette took about 500 lines of code and markup, not including the jQuery library. Much of this was needed to translate arbitrary CSS color strings into RGB values, rather than for user interface logic.

To allow users to test the behavior of the regular expression that they have entered, two columns are available below the expression input area. The left column contains an input area labeled ``Should match...'' and the right column contains an input area labeled ``Should NOT match...''. Users use these to enter lists of test  strings into each column. The background colors behind these strings change to indicate whether the regular expression that has been entered matches or does not match that string. Green indicates that the pattern  {\it matched} the string and red indicates that it does {\it not} match, regardless of the column that the test is in (this scheme was chosen based on feedback from an initial pilot of this palette, the results of which are not included below.) A key describing this color scheme is displayed after the first test has been entered (not shown above, see video).
